---
title: "Reproducible Research: Peer Assessment 1"
author: "Grace"
date: "Sunday, February 15, 2015"
output: html_document
html_document:
    keep_md: true 
---


## Loading and preprocessing the data
```{r}
setwd("C:/Data_wd/RR/coursework1_submission")
unzip("activity.zip")
data = read.csv("activity.csv",colClasses=c("integer","Date","integer"))
head(data)
library(ggplot2)

datanoNA = na.omit(data)
head(datanoNA)

```

## What is mean total number of steps taken per day?

1. Make a histogram of total number of steps taken for each day

```{r}
dataagg = aggregate(steps ~ date, data=datanoNA, FUN="sum")
head(dataagg)

ggplot(dataagg, aes(date,steps)) + 
  labs(title="Total number of steps taken each day",x="Date", y="Total number of steps") +
  geom_bar(stat = "identity", color="purple", fill="purple")
```

2. Calculate and report the mean and median total number of steps taken per day

Mean steps taken per day:
```{r}
meanstepsdf = aggregate(steps~date,data=datanoNA, FUN="mean")
head(meanstepsdf)
```

Median steps taken per day:
```{r}
medianstepsdf = aggregate(steps~date,data=datanoNA, FUN="median")
head(medianstepsdf)
```

Mean steps for all days:
```{r}
o_mean=mean(dataagg$steps)
o_mean
```

Median steps for all days:

```{r}
o_median=median(dataagg$steps)
o_median
```

## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
avgSteps <- aggregate(steps~interval, data=datanoNA, FUN="mean")
head(avgSteps)
names(avgSteps)[2] <- "meanSteps"
head(avgSteps)

ggplot(avgSteps, aes(interval, meanSteps)) + geom_line(color = "purple", size = 1.0) + labs(title = "Time series plot of 5-min interval of average steps taken, \n averaged across all days", x = "5-minute intervals", y = "Average number of steps taken")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
avgSteps[avgSteps$meanSteps == max(avgSteps$meanSteps), ]$interval
```

## Imputing missing values
1. The total number of missing values in the dataset:

```{r}
sum(is.na(data))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Strategy: To use the mean steps of the 5-minute interval to fill each NA in steps column if interval matches.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
newData <- data 
for (i in 1:nrow(newData)) {
    if (is.na(newData$steps[i])) {
        newData$steps[i] <- avgSteps[which(newData$interval[i] == avgSteps$interval), ]$meanSteps
    }
}

head(newData)
sum(is.na(newData))
```


4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r}

dataagg2 = aggregate(steps~date, data=newData, FUN="sum")
head(dataagg2)

ggplot(dataagg2, aes(date, steps)) + 
  labs(title = "Total Number of steps taken each day (imputed data)", x = "Date", y = "Total number of steps") +
  geom_bar(stat = "identity", color="purple", fill="purple") 
```

5. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

Mean steps taken per day:
```{r}
meanstepsdf2 = aggregate(steps~date,data=newData, FUN="mean")
head(meanstepsdf2)
```

Median steps taken per day:
```{r}
medianstepsdf2 = aggregate(steps~date,data=newData, FUN="median")
head(medianstepsdf2)
```

Mean steps for all days:
```{r}
newmean = mean(dataagg2$steps)
newmean
```

Median steps for all days:
```{r}
newmedian = median(dataagg2$steps)
newmedian
```

Compare them with the two before imputing missing data:
```{r}
newmean - o_mean
newmedian - o_median
```

After imputing the missing data, the new mean of total steps taken for all days is the same as that of the old mean while the new median of total steps taken for all days is slightly greater than that of the old median.


## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
head(newData)
newData$day = factor(format(newData$date, "%A"))
newData$day_end = factor(format(newData$date, "%A"))
levels(newData$day_end)
levels(newData$day_end) <- list(weekday = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),
                                 weekend = c("Saturday", "Sunday"))
levels(newData$day_end)
table(newData$day_end)
head(newData)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
newavgSteps = aggregate(steps~interval+day_end, data=newData, FUN="mean") 
head(newavgSteps)
names(newavgSteps)[3] <- "meanSteps"
head(newavgSteps)

library(lattice)
xyplot(meanSteps ~ interval | day_end, newavgSteps, 
       layout = c(1, 2), type = "l", 
       xlab = "Interval", ylab = "Mean Number of steps")
```

